A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-21 manuscript.tex
:construction: R sections added
:construction: input variables cleaned up

Recently, one promising way of identifying person-level developmental patterns has been \textit{time series clustering} --- the idea of inductively grouping participants based on similarities of their time series \citep{ariens2020}. This analysis type essentially seeks to capture comparable within-person developments --- often with some form of numeric summary --- and then groups the persons based on the extracted summaries \citep[][]{liao2005}\footnote{It should be noted in some cases the time series do not need to be summarized and can be compared directly. Such analyses are however only possible with highly regular and controlled data, such as EEG data \citep{huang1985}, or when clustering variables within the person rather than identifying differences between persons \citep{haslbeck2022}. Multivariate data from extensive psychological survey studies, as we discuss here, are seldomly directly comparable across persons \citep[e.g., ][]{faloutsos1994}. We provide a broader embedding of the current methods for psychological time series in the discussion section.}. Time series clustering, thus, crucially depends on identifying meaningful summaries of the time series developments, which can be used to compare participants \citep[][]{Aghabozorgi2015}.

@@

Yet, despite this rich diversity of meaningful time series features in psychology, most clustering of ESM data has only focused on a small and restrictive selection of time series characteristics. Thus far, the most common approach has been to cluster participants based on person-specific model parameters --- notably intercepts and slopes from vector autoregression models \citep[VAR; e.g.,][]{ariens2020, bulteel2016, stefanovic2022}. While such model parameters are familiar to researchers in the field, they are often restricted to autocorrelations and cross-lagged partial correlations \citep[e.g.,][]{bringmann2018c} --- only two of the many potentially important time series features. Additionally, the validity of model parameters is traditionally restricted by the assumptions of the particular statistical model. To take the common VAR model as an example, the model explicitly assumes that the time intervals between measurements are consistent (i.e., equidistant measurement assumption), it does not allow for missing values, and it assumes that means or variances do not change over time \citep[i.e., stationarity assumption;][]{lutkepohl2005}. All of these violated assumptions reflect types of data researchers commonly collect to address important questions of erratic, and context-specific phenomena \citep[][]{myin-germeys2018, hamaker2017, kivela2022, helmich2020a}. Model parameters might, thus, not always accurately capture the time series and are often restricted in the features they capture.

@@

Recent efforts to address the shortcomings of model parameters for time series clustering have primarily proposed more complicated models, which either seek to relax specific assumptions \citep[e.g.,][]{denteuling2021} or include additional time series features \citep[e.g., see][]{krone2018}. In this manuscript, we instead, propose to directly cluster based on important and relevant features. Adequately named \textit{feature-based time series clustering}, such an approach has been a common procedure in digital phenotyping \citep[][]{loftus2022} and the broader machine learning literature \citep[][]{maharaj2019}. As such, the analysis has been applied to a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. We argue that for psychological time series data, feature-based clustering offers a flexible variety, theory-based selection, fewer strict assumptions, easy and intuitive analysis, and meaningful psychological interpretability.

@@

To illustrate the functioning and utility of feature-based time series clustering with psychological ESM data, we apply the clustering process to a recent set of studies that collected data from research on migration experiences. Researchers have recently started using ESM data to follow the daily interactions of migrants with the cultural majority groups \citep[e.g.,][]{Keil2020} and this type of research exemplifies some of the real-world data issues that ESM data commonly faces. The new type of research comes as a response to prominent reviews within the migration literature, which have called for more longitudinal \citep[e.g.,][]{Ward2019} and real-world data \citep[e.g.,][]{McKeown2017}, while at the same time, conceptual works have pushed for more diverse assessments of migration experiences, including motivational, affective, cognitive, and behavioral aspects \citep[e.g.,][]{Kreienkamp2022d}. The migration ESM research, thus, offers a good example of the ESM data that introduces a multivariate conceptualization with event-specific missingness patterns.

[...]

@@

Introduction on the structure ...

Armed with a selection of potentially meaningful and accurate time series features to address a given research question, we can consider the practical steps of feature-based time series clustering. The feature-based clustering analysis can be structured into four main steps \citep{rasanen2009, wang2006}. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, with an optional feature reduction step if there are too many data points for the clustering algorithms, (3) the actual clustering of the time series features, and (4) the evaluation and interpretation of the clusters. It is important to note that these steps are a general outline, and the specific details of the analysis will depend on the nature of the data and the research question being addressed. Nonetheless, below we will briefly describe each of the steps conceptually. We also provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

\paragraph{Input variables}
Time series clustering starts with the selection and preparation of the variables of interest. While the selection will necessarily be field- and concept-specific, there are a few conceptual and methodological issues that should be considered. Conceptually, the included variables should adequately capture the concept of interest and should be meaningful to the understanding of the time series. There are, for example, calls that emotion dynamics should be assessed with a repertoire of positive and negative emotions \citep[e.g.,][]{dejonckheere2019}, migration experiences are fully captured with affect, behavior, cognition, and desire measurements \citep[e.g.,][]{Kreienkamp2022d}, and many health developments are commonly captured within the biopsychosocial domains \citep[e.g.,][]{suls2004}. At the same time, however, the added number of variables can become a methodological concern. Not only can redundant and irrelevant variables diminish the quality of the analyses, but with intensive longitudinal data the number of data points compounds across participants, measurement occasions, and variables so that additional variables can make many of the following steps substantially more difficult (also see \fgrref{fig:TSCFlowN}).

@@

Once the important variables have been selected, the data needs to be prepared for the analysis steps. Importantly, this not only means validating and cleaning the data (e.g., re-coding, removing duplicate or unwanted measurements) but also making the time series comparable. Two important steps are making the time-frames and response scales comparable across participants --- for example, by choosing a time frame that is common to most participants and standardizing the participants' responses \citep{liao2005}.

\paragraph{Feature extraction}
The main feature extraction step ultimately depends on the choice of the selected time series features but generally speaking the process is comparatively simple. For each of the included variables the features are extracted for each participant and replace the original time series \citep{wang2006}\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

@@

\paragraph{Feature reduction}
Once a meaningful selection of time series features has been extracted for each variable and participant, the total number of data points sometimes remains too large for the desired clustering algorithm. As an example, a relatively common scenario would include 10 variables of interest, where 8 time series features are extracted, resulting in 80 features per participant (with a common sample size of 100 participants that would result in a total of 8,000 data points in this hypothetical example). We offer an illustration of the compounding numbers of data points in \fgrref{fig:TSCFlowN}. The difficulty of working with such a large number of dimensions is sometimes termed the `dimensionality curse' \citep[e.g.,][]{altman2018}. To deal with this dimensionality issue, two main approaches have been proposed --- feature selection and feature projection \citep[e.g.,][also see \tblref{tab:featureReduction} for a full overview of the approaches, methods, and common algorithms]{erdogmus2008}.

@@

\paragraph{Feature clustering}
For feature-based time-series clustering, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible \citep{liao2005}. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment \citep[or similarity measurement for qualitative features; see][]{Aghabozorgi2015}. While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, we will briefly introduce some of the more readily available approaches suitable for most time series feature data. The well-established and readily available clustering approaches can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof \citep[see \tblref{tab:clusterApproaches} for an overview; also see][for a broader review]{jain1999}.

@@

\paragraph{Cluster Evaluation}
The final evaluation step includes two main elements, (1) evaluating the performance of the clustering analysis and (2) interpreting the extracted clusters conceptually.

@@

% Methods and Results from RMarkdown render
\input{methods-and-results}
%\input{tables/descrWide} 
%\input{tables/descrLong}

>>>>

Recently, one promising way of identifying person-level developmental patterns has been \textit{time series clustering} --- the idea of inductively grouping participants based on similarities of their time series \citep{ariens2020}. This analysis type essentially seeks to capture comparable within-person developments --- such as whether a variable remains stable over time, consistently increases, or exhibits cyclical patterns --- and then groups the persons based on these patterns  \citep[][]{liao2005}\footnote{It should be noted in some cases the time series do not need to be summarized and can be compared directly. Such analyses are however only possible with highly regular and controlled data, such as EEG data \citep{huang1985}, or when clustering variables within the person rather than identifying differences between persons \citep{haslbeck2022}. Multivariate data from extensive psychological survey studies, as we discuss here, are seldomly directly comparable across persons \citep[e.g., ][]{faloutsos1994}. We provide a broader embedding of the current methods for psychological time series in the discussion section.}. Time series clustering, thus, crucially depends on identifying meaningful summaries of the time series developments, which can be used to compare participants \citep[][]{Aghabozorgi2015}.

@@

Yet, despite this rich diversity of meaningful time series features in psychology, most clustering of ESM data has only focused on a small and restrictive selection of time series characteristics. Thus far, the most common approach has been to cluster participants based on person-specific model parameters --- notably intercepts and slopes from vector autoregression models \citep[VAR; e.g.,][]{ariens2020, bulteel2016, stefanovic2022}. While such model parameters are familiar to researchers in the field, they are often restricted to autocorrelations and cross-lagged partial correlations \citep[e.g.,][]{bringmann2018c} --- only two of the many potentially important time series features. Additionally, the validity of model parameters is traditionally restricted by the assumptions of the particular statistical model. To take the common VAR model as an example, the model explicitly assumes that the time intervals between measurements are consistent (i.e., equidistant measurement assumption), it does not allow for missing values, and it assumes that means or variances do not change over time \citep[i.e., stationarity assumption;][]{lutkepohl2005}. These assumptions however stand incontrast to types of data researchers commonly collect to address important questions of erratic, and context-specific phenomena \citep[][]{myin-germeys2018, hamaker2017, kivela2022, helmich2020a}. Model parameters might, thus, not always accurately capture the time series and are often restricted in the features they capture.

@@

Recent efforts to address the shortcomings of model parameters for time series clustering have primarily proposed more complicated models, which either seek to relax specific assumptions \citep[e.g.,][]{denteuling2021} or include additional time series features \citep[e.g., see][]{krone2018}. In this manuscript, we instead, propose to directly cluster based on important and relevant features. Adequately named \textit{feature-based time series clustering}, such an approach has been a common procedure in digital phenotyping \citep[][]{loftus2022} and the broader machine learning literature \citep[][]{maharaj2019}. As such, the analysis has been applied to a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. ESM APPLICATION. We argue that for psychological time series data, feature-based clustering offers a flexible variety, theory-based selection, fewer strict assumptions, easy and intuitive analysis, as well as meaningful psychological interpretability.

@@

To illustrate the functioning and utility of feature-based time series clustering with psychological ESM data, we apply the clustering process to a recent set of studies that collected data from research on migration experiences. Researchers have recently started using ESM data to follow the daily interactions of migrants with the cultural majority groups \citep[e.g.,][]{Keil2020}. This type of research comes as a response to prominent reviews within the migration literature, which have called for more longitudinal \citep[e.g.,][]{Ward2019} and real-world data \citep[e.g.,][]{McKeown2017}, while at the same time, conceptual works have pushed for more diverse assessments of migration experiences, including motivational, affective, cognitive, and behavioral aspects \citep[e.g.,][]{Kreienkamp2022d}. Importantly for our illustration, the migration ESM research, thus, exemplifies some of the real-world data issues that ESM data commonly faces, including a multivariate conceptualization with event-specific missingness patterns.

\input{latex-results/01DataDescription.tex}

@@

To introduce and illustrate the feature-based clustering analysis, we will follow the conceptual steps of the procedure in sequential order and discuss key issues for each step. To do so we will follow the common separation that has structured feature-based clustering into four main steps \citep{rasanen2009, wang2006}. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, with an optional feature reduction step if there are too many data points for the clustering algorithms, (3) the actual clustering of the time series features, and (4) the evaluation and interpretation of the clusters. 
While this is a common conceptual separation of procedural elements, it is important to note that these steps are a general outline, and the specific details of the analysis will depend on the nature of the data and the research question being addressed. Nonetheless, the conceptual nature of these steps allows us to introduce the mayor elements of the analysis. We also provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

\subsection{Input variables}
Time series clustering starts with the selection and preparation of the variables of interest. While the selection will necessarily be field- and concept-specific, there are a few conceptual and methodological issues that should be considered. Conceptually, the included variables should adequately capture the concept of interest and should be meaningful to the understanding of the time series. One of the advantages of feature-based clustering is that it is inherently adept to accomodating multi-variate concepts --- a common aim in ESM research. There are, for example, calls that emotion dynamics should be assessed with a repertoire of positive and negative emotions \citep[e.g.,][]{dejonckheere2019}, many health developments are captured within the biopsychosocial domains \citep[e.g.,][]{suls2004}, and migration experiences are fully captured with affect, behavior, cognition, and desire measurements \citep[e.g.,][]{Kreienkamp2022d}. At the same time, however, the added number of variables can become a methodological concern. Not only can redundant and irrelevant variables diminish the quality of the analyses, but with intensive longitudinal data the number of data points compounds across participants, measurement occasions, and variables so that additional variables can make many of the following steps substantially more difficult (also see \fgrref{fig:TSCFlowN}).

@@

\input{latex-results/02InputVariables.tex}

\subsection{Feature extraction}
The main feature extraction step ultimately depends on the choice of the selected time series features but generally speaking the process is comparatively simple. For each of the included variables the features are extracted for each participant and replace the original time series \citep{wang2006}\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

@@

\input{latex-results/03FeatureExtraction.tex}

\subsubsection{Feature reduction}
Once a meaningful selection of time series features has been extracted for each variable and participant, the total number of data points sometimes remains too large for the desired clustering algorithm. As an example, a relatively common scenario would include 10 variables of interest, where 8 time series features are extracted, resulting in 80 features per participant (with a common sample size of 100 participants that would result in a total of 8,000 data points in this hypothetical example). We offer an illustration of the compounding numbers of data points in \fgrref{fig:TSCFlowN}. The difficulty of working with such a large number of dimensions is sometimes termed the `dimensionality curse' \citep[e.g.,][]{altman2018}. To deal with this dimensionality issue, two main approaches have been proposed --- feature selection and feature projection \citep[e.g.,][also see \tblref{tab:featureReduction} for a full overview of the approaches, methods, and common algorithms]{erdogmus2008}.

@@

\input{latex-results/04FeatureReduction.tex}

\subsection{Feature clustering}
For feature-based time-series clustering, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible \citep{liao2005}. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment \citep[or similarity measurement for qualitative features; see][]{Aghabozorgi2015}. While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, we will briefly introduce some of the more readily available approaches suitable for most time series feature data. The well-established and readily available clustering approaches can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof \citep[see \tblref{tab:clusterApproaches} for an overview; also see][for a broader review]{jain1999}.

@@

\input{latex-results/05FeatureClustering.tex}

\subsection{Cluster Evaluation}
The final evaluation step includes two main elements, (1) evaluating the performance of the clustering analysis and (2) interpreting the extracted clusters conceptually.

@@

\input{latex-results/06ClusterEvaluation.tex}
